# Developing a custom visualization

```{r, include=FALSE}
## Utility functions: show_palette()
source("scripts/utils.R")
knitr::opts_knit$set(cache = TRUE,
                     tidy = TRUE,
                     tidy.opts = list(blank = FALSE, width.cutoff = 60))
library(ggplot2)
library(tidyr)
```

## An experimental PTSD treatment 

This chapter shows the iterative process of building a visualization where both the audience and the data story are taken into consideration.

The _story_ revolves around data that was collected in a research effort investigating the effect of some treatment of subjects with PTSD (Post-traumatic stress disorder). Only one variable of that dataset is shown here, a stress score.

Since the group size was very low, and there was no control group, statistical analysis was not really feasible.

But the question was: is there an indication of positive effect and a reason to continue the investigations?
An attempt was made by developing a visualization to answer this question.

### The data {-}

The collected data was a distress score collected at three time points: 0 months (null measure, T0), 3 months (T1) and 12 months (T2) through questionnaires.

```{r read-distress-data, echo = FALSE, out.width="80%"}
distress_data <- read.table("data/distress_score_timeseries.txt", sep = "\t", header = T, dec=",")
knitr::kable(distress_data, caption = "Distress data")
```

You can see this is a really small dataset. 

### Choose a visualization {-}

Before starting the visualization, several aspects should be considered:

- The **audience**:
    - people do not want to read lots of numbers in a table\
    - in this case no knowledge of statistics (and this is usually the case)
- The **data**:
    - here, small sample size is an issue
    - this dataset has connected measurements (timeseries-like)

For this dataset I chose a **_jitterplot_** as basis because it is well suited for small samples. A boxplot tends to be indicative of information that simply is not there with small datasets. Moreover, a boxplot has a certain complexity that people who are not schooled in statistics have problems with.

### Tidy the data {-}

To work with ggplot2, a tidy ("long") version of the data is required. 
In the next chapter this will be dealt with in detail. 
Here the T0, T1 and T2 columns are **_gather_**ed into a single column because they actually represent a single variable: Time. All measured stress values, also a single variable, are gathered into a single column as well. This causes a flattening of the data (less columns, more rows). 

```{r ptsd-tidy}
distress_data_tidy <- gather(distress_data,
                        key=Timepoint,
                        value=Stress, "T0", "T1", "T2")
distress_data_tidy$Timepoint <- factor(distress_data_tidy$Timepoint, ordered = T)
knitr::kable(head(distress_data_tidy, n = 10), caption = "Tidied data")
```

### A first version {-}

This is the first version of the cisualization. The jitter has been created with `geom_jitter`. The plot symbols have been made transparent to keep overlapping points visible. The plot symbols have been made bigger to support embedding in (powerpoint) presentations. A little horizontal jitter was introduced to have less overlap of the symbols, but not too much - the discrete timepoints still stand out well. Vertical jitter omitted since the data are already measured in a continuous scale. A typical use case for vertical jitter is when you have discrete (and few) y-axis measurements.

```{r ptsd-jitter-1, fig.asp=.75, out.width='80%', fig.align='center', fig.cap="A first attempt"}
ggplot(distress_data_tidy, aes(x=Timepoint, y=Stress)) +
    geom_jitter(width = 0.1, size = 2, alpha = 0.6)
```


### Add mean and SD {-}

To emphasise the trend in the timeseries, means and standard deviations from the mean were added using `stat_summary()`. Always be aware of the orders of layers of your plot! Here, the `stat_summary` was placed "below" the plot symbols. Again, size was increased for enhaced visibility in presentations. Why not the median? Because of the audience! Everybody knows what a mean is, but few know what a median is - especially at management level.

```{r ptsd-jitter-2, fig.asp=.75, out.width='80%', fig.align='center', fig.cap="With mean and standard deviation"}
mean.sd <- function(x) {
  c(y = mean(x), ymin = (mean(x) - sd(x)), ymax = (mean(x) + sd(x)))
}

ggplot(distress_data_tidy, aes(x = Timepoint, y = Stress)) +
    stat_summary(fun.data = mean.sd, color = "darkred", size = 1.5) +
    geom_jitter(width = 0.1, size = 2, alpha = 0.6)
```

### Emphasise worst cases {-}

To emphasise the development of subjects who were in the worst shape at the onset of the research (T0), the top 25% with respect to distress score at T0 were highlighted.

```{r ptsd-highlight-worst-at-T0, fig.asp=.75, out.width='80%', fig.align='center', fig.cap="Worst cases emphasised"}
distress_data$high_at_T0 <- ifelse(distress_data$T0 > quantile(distress_data$T0, 0.75), "Q4", "Q1-Q3")

distress_data_tidy <- gather(distress_data,
                        key=Timepoint,
                        value=Stress, "T0", "T1", "T2")
distress_data_tidy$Timepoint <- factor(distress_data_tidy$Timepoint, ordered = T)
knitr::kable(head(distress_data))
```

The color is added using `aes(color = high_at_T0)` within the `geom_jitter()` call.

```{r ptsd-jitter-3, fig.width=8, fig.height=4.5}
p <- ggplot(distress_data_tidy, aes(x=Timepoint, y=Stress)) +
    stat_summary(fun.data=mean.sd, color = "darkred", size = 1.5) +
    geom_jitter(width = 0.1, size = 2, alpha = 0.6, aes(color = high_at_T0))
p
```


## Last tweaks: fonts and legend

The plot is more or less ready. Now is the time to adjust the plot "theme".

```{r ptsd-jitter-5, fig.width=8, fig.height=4.5}
p + theme_minimal(base_size = 14) +
    theme(legend.position = "top") +
    labs(color="Group")
```

## The code 

Here is the code used for data preparation:

```{r ptsd-data-prep-code, eval=FALSE}

distress_data$high_at_T0 <- ifelse(
    distress_data$T0 > quantile(distress_data$T0, 0.75), "Q4", "Q1-Q3")

distress_data_tidy <- gather(distress_data,
                        key=Timepoint,
                        value=Stress, "T0", "T1", "T2")
distress_data_tidy$Timepoint <- factor(distress_data_tidy$Timepoint,
                                       ordered = T)

mean.sd <- function(x) {
  c(y = mean(x), ymin=(mean(x)-sd(x)), ymax=(mean(x)+sd(x)))
}
```


This is the final code for the plot

```{r ptsd-plot-code, eval=FALSE}
ggplot(distress_data_tidy, aes(x=Timepoint, y=Stress)) +
    stat_summary(fun.data=mean.sd, color = "darkred", size = 1.5) +
    geom_jitter(width = 0.1,
                size = 2,
                alpha = 0.6,
                aes(color = high_at_T0)) +
    labs(color="Group") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "top") +
    labs(color="Group")
```


