# Exercise Solutions {#exercise-solutions}

```{r, include=FALSE}
options(digits = 3)
knitr::opts_knit$set(cache = TRUE,
                     tidy = TRUE,
                     tidy.opts = list(blank = FALSE, width.cutoff = 60))
```


## Introduction

Only solutions chunks are displayed here, not the output - except for the plots.

## The `ggplot2` and `tidyr` packages

```{r message = FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(lubridate)
```


### Trees

**A [&#10059;]**  

```{r trees-A, fig.asp=0.7, out.width='70%', fig.align='center', eval = FALSE}
ggplot(data = trees, 
       mapping = aes(x = Girth, y = Height)) +
    geom_point(color = "blue") +
    xlab('Girth (in)') +
    ylab('Height (ft)')
```

**B [&#10059;&#10059;]**  

```{r trees-B, fig.asp=0.7, out.width='80%', fig.align='center', eval = FALSE}
ggplot(data = trees, 
       mapping = aes(x = Girth, y = Height)) +
    geom_point(color = "blue") +
    geom_smooth(method = "lm", se = FALSE) +
    xlab('Girth (in)') +
    ylab('Height (ft)')
```

**C [&#10059;&#10059;]**  

Final figure.

```{r trees-C, fig.asp=0.7, out.width='80%', fig.align='center', fig.cap="Tree girth, height and volume relationship"}
ggplot(data = trees, 
       mapping = aes(x = Girth, y = Height)) +
    geom_point(aes(size = Volume), color = "blue", alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE) +
    xlab('Girth (in)') +
    ylab('Height (ft)') +
    labs(size = 'Volume (ft^3)')
```

### Insect Sprays 

The `datasets` package that is shipped with R has a dataset called `?`. Type `?InsectSprays` to get information on it. 

**A [&#10059;]**  

```{r insect-sprays-A, fig.asp=0.7, out.width='80%', fig.align='center', fig.cap = "The counts of insects in agricultural experimental units treated with different insecticides"}
ggplot(data = InsectSprays, 
       mapping = aes(x = spray, y = count)) +
    geom_boxplot() +
    ylab("Insect count") +
    xlab("Insect spray")
```

**B [&#10059;&#10059;]**  
Create a jitter plot, splitting on the spray type and jittering minimally on the y-axis. Make sure that overlapping plot symbols do not hide each other and try to find a nice color. 
[Extra: Give each spray a different plot symbol.]

```{r insect-sprays-B, fig.asp=0.7, out.width='80%', fig.align='center', fig.cap = "The counts of insects in agricultural experimental units treated with different insecticides"}
ggplot(data = InsectSprays, 
       mapping = aes(x = spray, y = count, color = spray)) +
    geom_jitter(height = 0, width = 0.1, shape = 18, size = 2, alpha = 0.7) +
    ylab("Insect count") +
    xlab("Insect spray")
```

### Diauxic growth 

**A [&#10059;&#10059;]**  

```{r diauxic-growth-A}
remote <- "https://raw.githubusercontent.com/MichielNoback/datasets/master/diauxic_growth/monod_diauxic_growth.csv"
local <- "../diauxic.csv"
#download.file(url = remote, destfile = local)
diauxic <- read.table(local, sep = ";", header = T)
diauxic <- pivot_longer(data = diauxic,
                        cols = -1,
                        names_to = "Substrate",
                        values_to = "OD")
```

**B [&#10059;]**  

```{r diauxic-growth-B}
diauxic$Substrate <- factor(diauxic$Substrate, 
       levels = c("GlucMann", "GlucXyl", "GlucArab", "GlucRham"), 
       labels = c("Glucose Mannose", "Glucose Xylose", "Glucose Arabinose", "Glucose Rhamnose"))
```

**C [&#10059;&#10059;]**  

Create a line plot with all four growth curves within a single graph.

```{r diauxic-growth-C, fig.asp=0.7, out.width='80%', fig.align='center', warning = FALSE, fig.cap = "Monod's Diauxic shift experiment."}
ggplot(data = diauxic,
       mapping = aes(x = Time, y = OD, color = Substrate)) +
    geom_point() +
    stat_smooth(method = "loess", se = FALSE, span = 0.3) +
    theme_bw()
```


**D [&#10059;&#10059;&#10059;]**  

```{r diauxic-growth-D, fig.asp=0.7, out.width='80%', fig.align='center', warning = FALSE, fig.cap = "Monod's Diauxic shift experiment."}
ggplot(data = diauxic,
       mapping = aes(x = Time, y = OD)) +
    geom_point() +
    stat_smooth(method = "loess", se = FALSE, span = 0.3) +
    facet_wrap(. ~ Substrate, nrow = 2) +
    theme_bw()
```


### Virginia Death Rates

```{r VADeaths-prep-rep}
library(dplyr)
## %>% is used to pipe results from one operation to the other, just like '|' in Linux.
virginia_death_rates <- as_tibble(VADeaths)
virginia_death_rates <- virginia_death_rates %>% 
    mutate("Age Group" = factor(rownames(virginia_death_rates), ordered = TRUE)) %>% 
    select(`Age Group`, everything()) #reorder the columns
```

**A (&#10059;&#10059;;&#10059;)**  

Pivot this table to long (tidy) format. This should generate a dataframe with four columns: `Age Group`, `Habitat`, `Gender` and `DeathRate`.

```{r VADeaths-A, results = 'hide'}
virginia_death_rates <- virginia_death_rates %>% pivot_longer(cols = -1, 
                 names_to = c("Habitat", "Gender"), 
                 names_sep = " ", 
                 values_to = "DeathRate")
```

**B (&#10059;&#10059;)**  


```{r VADeaths-B, fig.asp=0.7, out.width='70%', fig.align='center', fig.cap="Virginia Death rates"}
ggplot(data = virginia_death_rates, aes(Gender)) +
    geom_bar(aes(weight = DeathRate, fill = Habitat), position = "dodge")
```


### Investigate new visualization <sup><font size="3em">[&#10059;&#10059;]</font></sup>

This assignment has no solution of course. It is included here solely to keep the numbering consistent for both assignment chapters.


### ToothGrowth 
**[&#10059;&#10059;]**  

In my opinion, the boxplot is an excellent choice for this visualization. However, `geom_jitter()` could also work but is more of a hassle to split out on Dose.

```{r toothgrowth, fig.asp=0.7, out.width='80%', fig.align='center', fig.cap=" length of odontoblasts with vitamin C"}
ggplot(data = ToothGrowth, mapping = aes(x = supp, y = len)) +
    geom_boxplot(aes(color = as.factor(dose))) +
    xlab("Supplement") +
    ylab("Tooth length (mm)") +
    labs(color = "Dose (mg/day)")
```


### Puromycin 
**[&#10059;&#10059;&#10059;]**  

```{r puromycin2, fig.asp=0.7, out.width='80%', fig.align='center', fig.cap="reaction velocity versus substrate concentration"}
ggplot(data = Puromycin,
       mapping = aes(x = conc, y = rate, color = state)) +
    geom_point() +
    geom_smooth(method = "loess", formula = y ~ log(x), se = F)
```


### Global temperature

Load the data.

```{r}
remote_file <- "https://raw.githubusercontent.com/MichielNoback/datasets/master/global_temperature/annual.csv"
local_file <- "../annual.csv"

if (! file.exists(local_file)) {
    download.file(remote_file, destfile = local_file)
}

global_temp <- read.table("annual.csv",
                          header = TRUE,
                          sep = ",")
```

#### Create a scatter-and-line-plot [&#10059;&#10059;]

```{r scatter-global-temp, fig.asp=0.7, out.width='80%', fig.align='center', fig.cap="Global temperature anomalies"}
ggplot(data = global_temp,
       mapping = aes(x = Year, y = Mean, color = Source)) +
    geom_point(size = 0.5) +
    geom_line() +
    geom_smooth(se = FALSE, method = "loess") +
    theme_bw()
```

#### Re-create the heatmap [&#10059;&#10059;&#10059;]

```{r heatmap-global-temp, fig.asp=0.3, out.width='80%', fig.align='center', fig.cap="Global temperature anomalies"}
ggplot(data = global_temp[global_temp$Source == "GCAG", ],
       mapping = aes(x = Year, y = 1)) +
    geom_tile(aes(fill = Mean), colour = "white") + 
    scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
    theme_bw() +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.title.y = element_blank())
```

Note: rescaling the temperature from 0 to 1 may yield even better results.

### Epilepsy drug trial

#### Load data <sup><font size="3em">[&#10059;]</font></sup>

```{r results = 'hide'}
epilepsy_file <- "https://raw.githubusercontent.com/MichielNoback/datasets/master/epilepsy/epilepsy.csv"
epilepsy <- read.table(epilepsy_file,
                        header = TRUE,
                        sep = ",")
epilepsy$period <- factor(epilepsy$period, ordered=TRUE)
```

#### Reorganize the data <sup><font size="3em">[&#10059;&#10059;]</font></sup>

```{r results = 'hide'}
epilepsy <- as_tibble(
    epilepsy[, c("subject","age","base","treatment","period","seizure.rate")])
```

#### Create plots of seizure rates

**A [&#10059;&#10059;]**   

I used the `color` aesthetic here:

```{r boxplot-seizure-rates, fig.asp=0.7, out.width='80%', fig.align='center', fig.cap="Seizure rates with color"}
ggplot(data = epilepsy,
       mapping = aes(x = period, y = seizure.rate, color = treatment)) +
    geom_boxplot()
```

This is the alternative with `facet_wrap()`

```{r boxplot-seizure-rates2, fig.asp=0.7, out.width='80%', fig.align='center', fig.cap="Seizure rates with facet wrap"}
ggplot(data = epilepsy,
       mapping = aes(x = period, y = seizure.rate)) +
    geom_boxplot() +
    facet_wrap(. ~ treatment)
```

**B [&#10059;&#10059;&#10059;]**   

Again, this could be done with color or facetting, but in this case `shape` is also an option (not shown). I used a little trick to separate the `period` for treated/untreated.

```{r boxplot-seizure-rates3, fig.asp=0.7, out.width='80%', fig.align='center', fig.cap="Seizure rates"}
tmp <- epilepsy
tmp$period <- as.integer(tmp$period)
tmp$period_offset <- ifelse(epilepsy$treatment == "placebo", tmp$period - 0.10, tmp$period + 0.10)
#tmp
ggplot(data = tmp,
       mapping = aes(x = period_offset, y = seizure.rate, color = treatment)) +
    geom_jitter(width = 0.05, alpha = 0.7) 
```

**C [&#10059;&#10059;&#10059;]**   

```{r boxplot-seizure-rates4, fig.asp=0.7, out.width='90%', fig.align='center', fig.cap="Seizure rates"}
ggplot(data = epilepsy,
       mapping = aes(x = period, y = seizure.rate, color = treatment)) +
    geom_boxplot(aes(fill = treatment), alpha = 0.2) +
    geom_jitter(data = tmp, aes(x = period_offset), width = 0.05, height = 0, alpha = 0.5, shape = 18, size = 2) +
    theme_bw()
```

Note that it would have been nicer to omit the outliers from the boxplot. This is your challenge.

There are many more variations possible! You should really explore some of these as practice.

#### A boxplot after correction <sup><font size="3em">[&#10059;&#10059;&#10059;]</font></sup>

Correction:

```{r results = 'hide'}
epilepsy_corrected <- aggregate(
    seizure.rate ~ subject + age + base + treatment, 
    data = epilepsy, 
    FUN = sum)
epilepsy_corrected$seizure_rate_corrected <- epilepsy_corrected$seizure.rate - epilepsy_corrected$base
```

The plot:

```{r boxplot-seizure-rates5, fig.asp=0.7, out.width='80%', fig.align='center', fig.cap="Seizure rates after correction"}
ggplot(data = epilepsy_corrected,
       mapping = aes(x = treatment, y = seizure_rate_corrected)) +
    geom_boxplot()
```

#### Test for statistical significance <sup><font size="3em">[&#10059;&#10059;&#10059;]</font></sup>

```{r}
placebo <- epilepsy_corrected$seizure_rate_corrected[epilepsy_corrected$treatment == "placebo"]
treated <- epilepsy_corrected$seizure_rate_corrected[epilepsy_corrected$treatment == "Progabide"]
t.test(x = placebo, y = treated)

##better:
t.test(seizure_rate_corrected ~ treatment, data = epilepsy_corrected)
```

Note: you should actually test for normality sing the `shapiro.test()` function before embarking on a t-test analysis:

```{r results = 'hide'}
shapiro.test(treated)
shapiro.test(placebo)
```

So this is not the correct test; you should use the Mann-Whitney U test.

```{r}
wilcox.test(seizure_rate_corrected ~ treatment, data = epilepsy_corrected)
```

### Bacterial growth curves with Varioscan

#### Load, preprocess and tidy <sup><font size="3em">[&#10059;&#10059;&#10059;&#10059;]</font></sup>

Load:

```{r load-growth-data}
growth_data_file <- "data/varioscan/2020-01-30_wide.csv"
growth_data <- read.csv(growth_data_file)
```

Preprocess. Applying corrctions to the dataset is not easy as it is; this is much easier when the dataframe is transposed so samples are in columns and timepoints in rows. Also, transposition will force all data into character because the first columns contain character data. To prevent this, they are removed (and stored), leaving only the numerical OD measurements.

```{r preprocess-growth-data, eval = FALSE}
## Add rownames to get a hold of them after transpose
row_names <- paste0(growth_data$Content, '.',
                    growth_data$.copy, '.',
                    rep(LETTERS[1:8], times = 12)) #letters A-H represent the 8 dilutions
rownames(growth_data) <- row_names

## remove non-numeric cols but store them first
label_columns <- growth_data[, 1:4]
tmp <- growth_data[, -(1:4)]

## transpose
tmp <- t(tmp)

## build a new dataframe/tibble, empty but with predefined number of rows
growth_data_corr <- tibble(.rows = nrow(tmp))
```

Carry out the corrections; can this be done more efficiently? Take the challenge!

```{r correct-growth-data, eval = FALSE}
##1  9 17 25 33 41 49 57 65 73 81 89

for (col_select in 1:8) {
    #Using correction columns 25-32
    #Red.1: columns 1 to 8
    growth_data_corr[, row_names[col_select]] <- tmp[ , col_select] - tmp[ , col_select + 24]
    #Red.2: columns 9 to 16
    growth_data_corr[, row_names[col_select + 8]] <- tmp[ , col_select + 8] - tmp[ , col_select + 24]
    #Red.3: columns 17 to 24
    growth_data_corr[, row_names[col_select + 16]] <- tmp[, col_select + 16] - tmp[, col_select + 24]

    #Using correction columns 57-64
    #White.1: columns 33 to 40
    growth_data_corr[, row_names[col_select + 32]] <- tmp[ , col_select + 32] - tmp[ , col_select + 56]
    #White.2: columns 41 to 48
    growth_data_corr[, row_names[col_select + 40]] <- tmp[ , col_select + 40] - tmp[ , col_select + 56]
    #White.3: columns 49 to 56
    growth_data_corr[, row_names[col_select + 48]] <- tmp[ , col_select + 48] - tmp[ , col_select + 56]

    #Using correction columns 89-96
    #Elution.1: columns 65 to 72
    growth_data_corr[, row_names[col_select + 64]] <- tmp[ , col_select + 64] - tmp[ , col_select + 88]
    #Elution.2: columns 73 to 80
    growth_data_corr[, row_names[col_select + 72]] <- tmp[ , col_select + 72] - tmp[ , col_select + 88]
    #Elution.1: columns 81 to 89
    growth_data_corr[, row_names[col_select + 80]] <- tmp[ , col_select + 80] - tmp[ , col_select + 88]
}
growth_data_corr <- growth_data_corr[, row_names[-c(25:32, 57:64, 89:96)]]
```

Re-attach the 
```{r, eval = FALSE}
## Transpose back and re-attach the first columns and time series labels
growth_data_corr <- t(growth_data_corr)
label_columns <- label_columns[-c(25:32, 57:64, 89:96), ]
growth_data_corr <- cbind(label_columns, growth_data_corr)
names(growth_data_corr) <- c(colnames(label_columns), rownames(tmp))
```

Tidy:
```{r, eval = FALSE}
growth_data_tidy <- growth_data_corr %>%
    pivot_longer(cols = starts_with("T."),
             names_prefix = "T.",
             names_to = "Time",
             values_to = "OD") %>%
    mutate(Time = as.integer(Time)/60,
           Dilution = factor(Dilution, ordered = T),
           Copy = factor(.copy, ordered = T)) %>%
    select(Content, Copy, Dilution, Time, OD)
```

Note that using the `names_prefix = "T."` argument makes the `T.` prefix disappear so converting time to a number is easier. This conversion is done in the last step of this workflow, and it is also converted to hours instead of minutes. `Dilution` and `.copy` are converted to factors so they can be used as discrete values in ggplot2.


#### Create a growth curve visualization <sup><font size="3em">[&#10059;&#10059;&#10059;&#10059;]</font></sup>

Let's first look at the three copies of the same sample. Can they maybe be merged?

```{r growth-curve1, fig.asp=0.7, out.width='90%', fig.align='center', fig.cap="Duplicates only", eval = FALSE}
library(ggplot2)
growth_data_tidy %>%
    filter(Content == "Red", ) %>%
    ggplot(mapping = aes(x = Time, y = OD)) +
        geom_line(aes(color = Dilution, linetype = Copy)) +
    xlab("Time (H)")
```



```{r growth-curve2, fig.asp=0.7, out.width='90%', fig.align='center', fig.cap="Do anthocyans inhibit bacterial growth?", eval = FALSE}

growth_data %>% filter(.copy == 1 & Dilution %in% c(0, 0.25, 2)) %>%
    ggplot(mapping = aes(x = Time, y = OD)) +
        geom_line(aes(linetype = Dilution, color = Content)) +
    scale_color_manual()
```



### The dinos

#### export to csv <sup><font size="3em">[&#10059;&#10059;]</font></sup>

This exercise has no code. I exported the data and placed the csv files under `data/dinos/` of this repo.

#### clean up and load `codebook.csv`

Non-data lines were deleted. Lines 1-18 relate to skeleton data and 19-29 to footprint data. 

**A [&#10059;]**  

The separator is ": " but it is not possible to define multi-character separators in `read.table()`. For now, simply use ":".

```{r results = 'hide'}
codebook_file <- "data/dinos/codebook.csv"
codebook <- read.table(codebook_file, 
                      sep = ":",
                      stringsAsFactors = FALSE)
```

**B [&#10059;]**  

```{r results = 'hide'}
names(codebook) <- c('variable', 'description')
```


**C [&#10059;&#10059;]**  

```{r results = 'hide'}
codebook$description <- sub('^ ', '', codebook$description)
```

**D [&#10059;&#10059;]**  

```{r results = 'hide'}
codebook$dataset <- c(rep('skeleton', 18), rep('footprint', 11))
```

#### Write a utility function <sup><font size="3em">[&#10059;&#10059;&#10059;]</font></sup>

```{r}
get_description <- function(column_name, dataset = 'skeleton') {
    if (! is.element(dataset, codebook$dataset)) {
        stop(paste0('dataset ', dataset, ' does not exist'))
    }
    if (! is.element(column_name, codebook$variable)) {
        stop(paste0('column ', column_name, ' does not exist'))
    }
    ##return the value
    codebook[codebook$dataset == dataset & codebook$variable == column_name, "description"]
}
get_description('Dinosaur_gen', 'skeleton')
```

#### Load skeleton data <sup><font size="3em">[&#10059;]</font></sup>

```{r results = 'hide'}
skeleton_file <- "data/dinos/skeletons.csv"

(skeleton <- read.table(skeleton_file, 
                        sep = ";",
                        header = TRUE,
                        dec = ",",
                        as.is = 1))
```


#### Plot species versus time <sup><font size="3em">[&#10059;&#10059;&#10059;]</font></sup>

Note the use of the utility function previously created, `get_description()`.

```{r species-vs-time, fig.asp=0.7, out.width='80%', fig.align='center', fig.cap="A first visualization"}
ggplot(data = skeleton,
       mapping = aes(x = Midpoint, y = log2(Total_spec))) +
    geom_point(aes(color = Epoch)) + 
    geom_smooth(method = "loess") +
    xlab(get_description("Midpoint")) +
    ylab(paste0("log2 of ", get_description("Total_spec"))) +
    scale_x_reverse()
```

#### Reproducing the publication figure

**A [&#10059;&#10059;]**  

We have not yet dealt with `dplyr` so we'll use base R for this. Note that the column `Total_gen` already contains the sum of all tetrapods.

```{r}
skeleton$genera_proportion <- (skeleton$Dinosaur_gen / skeleton$Total_gen) * 100
skeleton$specimen_proportion <- (skeleton$Dinosaur_spec / skeleton$Total_spec) * 100
```

**B [&#10059;&#10059;&#10059;]**  

```{r results = 'hide'}
skeleton_proportions <- aggregate(
    cbind(genera_proportion, specimen_proportion) ~ Epoch + Stage + Midpoint, 
    data = skeleton,
    FUN = mean)
```

**C [&#10059;&#10059;&#10059;]**  

```{r results = 'hide'}
skeleton_proportions_long <- pivot_longer(data = skeleton_proportions, 
             names_to = "level", 
             values_to = "Proportion", 
             cols = ends_with("_proportion"))
```

**D [&#10059;&#10059;&#10059;&#10059;]**   


First it needs to be flattened for use in ggplot2:

```{r plot-reprodiction1, fig.asp=0.7, out.width='80%', fig.align='center', fig.cap="A reproduction"}
ggplot(data = skeleton_proportions_long,
       mapping = aes(x = Midpoint, y = Proportion, color = level)) +
    geom_point() +
    scale_x_reverse() +
    scale_color_manual(values = c("blue", "red")) +
    xlab("Geological age (Myr)") +
    annotate("point", x = 231, y = 1, shape = 23, size = 5, fill = "goldenrod1") +
    annotate("text", x = 231, y = -3, label = "CPE")
```

**E [&#10059;&#10059;&#10059;&#10059;&#10059;]**   


Actually, this is not that simple to do with base R. What you want is a function that calculates the mean over a time period or over a fixed sequence of numbers. We'll take the mean over a sequence of 3 numbers. This could be a possible solution, using a `for` loop.

```{r}
rows <- nrow(skeleton_proportions)
moving_average_spec <- numeric(rows)
moving_average_gen <- numeric(rows)
n <- 2
## reversed traversal!
for (i in rows:1) {
    if (i >= rows - n) {
        series <- i:rows
    } else {
        series <- i:(i+n) ## beware of the parentheses!
    }
    moving_average_gen[i] <- mean(skeleton_proportions$genera_prop[series])
    moving_average_spec[i] <- mean(skeleton_proportions$specimen_prop[series])
}
##attach to the proportions dataframe
skeleton_proportions$specimen_moving_average <- moving_average_spec
skeleton_proportions$genera_moving_average <- moving_average_gen

#skeleton_proportions <- skeleton_proportions[, c(1:4, 6, 5, 7)] 

##recreate the long form with pivot_longer
skeleton_prop_avg_long <- pivot_longer(data = skeleton_proportions,
             cols = matches("_"),
             names_pattern = "(genera|specimen)_",
             values_to = c("proportion", "moving_average"),
             names_to = "level")
```

Finally, plot:

```{r plot-reprodiction2, fig.asp=0.7, out.width='80%', fig.align='center', fig.cap="A reproduction"}

ggplot(data = skeleton_prop_avg_long) +
    geom_point(mapping = aes(x = Midpoint, y = proportion, color = level)) +
    geom_line(mapping = aes(x = Midpoint, y = moving_average, color = level)) +
    scale_x_reverse() +
    scale_color_manual(values = c("blue", "red")) +
    xlab("Geological age (Myr)") +
    annotate("point", x = 231, y = 1, shape = 23, size = 5, fill = "goldenrod1") +
    annotate("text", x = 231, y = -3, label = "CPE")
```

**[THIS IS NOT CORRECT YET!!]**


#### Make a better figure <sup><font size="3em">[&#10059;&#10059;&#10059;&#10059;]</font></sup>

The Loess regression curve is a favorite of mine.

```{r plot-reprodiction3, fig.asp=0.7, out.width='80%', fig.align='center', fig.cap="A reproduction"}
ggplot(data = skeleton_prop_avg_long, 
       mapping = aes(x = Midpoint, y = proportion, color = level)) +
    geom_point() +
    geom_smooth(method = "loess", se = FALSE) +
    scale_x_reverse() +
    scale_color_manual(values = c("blue", "red")) +
    xlab("Geological age (Myr)") +
    annotate("point", x = 231, y = 1, shape = 23, size = 5, fill = "goldenrod1") +
    annotate("text", x = 231, y = -3, label = "CPE")
```

#### Use the size aesthetic <sup><font size="3em">[&#10059;&#10059;&#10059;&#10059;&#10059;]</font></sup>

We need more data for this, the totals for each "Stage":

```{r}
## re-create mean proportions
skeleton_proportions <- aggregate(
    cbind(genera_proportion, specimen_proportion) ~ Epoch + Stage + Midpoint, 
    data = skeleton,
    FUN = mean)

## first sum the finds:
skeleton_totals <- aggregate(
    cbind(Total_gen, Total_spec) ~ Midpoint,
    data = skeleton,
    FUN = sum)
tmp <- names(skeleton_totals)
tmp[2] <- "genera_totals"
tmp[3] <- "specimen_totals"
names(skeleton_totals) <- tmp

## combine with proportions
skeleton_prop_and_totals <- merge(skeleton_proportions, skeleton_totals, by = "Midpoint")

## reshuffling is required for pivot_longer to work as expected!!!
skeleton_prop_and_totals <- skeleton_prop_and_totals[, c(1:4, 6, 5, 7)]

## re-create the long form with pivot_longer
(skeleton_prop_totals_long <- pivot_longer(
    data = skeleton_prop_and_totals,
    cols = matches("_"),
    names_pattern = "(genera|specimen)_",
    values_to = c("proportion", "totals"),
    names_to = "level"))
```

Now for the plot

```{r plot-reprodiction4, fig.asp=0.7, out.width='80%', fig.align='center', fig.cap="A reproduction"}
ggplot(data = skeleton_prop_totals_long, 
       mapping = aes(x = Midpoint, y = proportion, color = level, size = totals)) +
    geom_point() +
    geom_smooth(method = "loess", se = FALSE) +
    scale_x_reverse() +
    scale_color_manual(values = c("blue", "red")) +
    xlab("Geological age (Myr)") +
    annotate("point", x = 231, y = 1, shape = 23, size = 5, fill = "goldenrod1") +
    annotate("text", x = 231, y = -3, label = "CPE")
```


## The `tidyr` and `dplyr` packages

### Global temperature revisited

**A [&#10059;&#10059;]**  

```{r pre-1970-anomaly, eval = FALSE}
global_temp %>% filter(Source == "GCAG" & Year < 1970 & Mean > 0.1)
global_temp %>% filter(Source == "GISTEMP" & Year < 1970 & Mean > 0.1)
```


**B [&#10059;&#10059;]**  

```{r coldest-post-1945, eval = FALSE}
global_temp %>% filter(Source == "GCAG" & Year > 1945) %>% 
    filter(Mean == min(Mean))

global_temp %>% filter(Source == "GISTEMP" & Year > 1945) %>% 
    filter(Mean == min(Mean))
```


**C [&#10059;&#10059;&#10059;]**  

This is a case for `filter(row_number() ...)` or `slice()`:

```{r warmest-five-years, eval = FALSE}
global_temp %>% 
    group_by(Source) %>% 
    arrange(desc(Mean)) %>% 
    slice(1:5) %>%
    #filter(row_number() %in% 1:5) %>% #works as well
    ungroup()
```


**D [&#10059;&#10059;&#10059;]**  

This is a case for splitting with `cut()`. Package dplyr has an equivalent named `ntile()`. 

```{r temperature-per-decade, eval = FALSE}

cut(global_temp$Year, breaks = seq(min(global_temp$Year), max(global_temp$Year), 10))

global_temp %>%
    filter(Year >= 1970) %>%
    mutate(Decade = cut(Year, 
                        breaks = seq(1970, 2020, 10), 
                        labels = seq(1970, 2010, 10),
                        right = FALSE)) %>%
    group_by(Decade, Source) %>%
    summarize(Mean = mean(Mean))
```


### ChickWeight

1. **Starting weight [&#10059;&#10059;]**.

```{r chick-starting-weight, eval = FALSE}
ChickWeight %>% 
    filter(Time == 0) %>%
    arrange(desc(weight)) %>%
    select(Chick, Diet, weight) %>%
    head(1)
```

2. **Total weight gain [&#10059;&#10059;&#10059;]**.

```{r chick-weight-gain, eval = FALSE}
ChickWeight %>% 
    group_by(Chick) %>%
    mutate(weight_gain = weight - lag(weight)) %>%
    summarize(timepoints = n(), 
              total_weight_gain = sum(weight_gain, na.rm = T),
              Diet = first(Diet)) %>%
     arrange(desc(total_weight_gain)) %>% head(4)
```

Note: you could also have selected the first and last measurement for each chic of course, but not simply the last timepoint since several chickens had died by then.

3. **Average weight gain per diet [&#10059;&#10059;&#10059;&#10059;]**. 

```{r chick-diet-weight-gain, eval = FALSE}
ChickWeight %>% 
    group_by(Chick) %>%
    mutate(weight_gain = weight - lag(weight)) %>%
    summarize(timepoints = n(), 
              total_weight_gain = sum(weight_gain, na.rm = T),
              Diet = unique(Diet)) %>%
    ungroup() %>%
    group_by(Diet) %>%
    summarize(num_chicks = n(),
              average_weight_gain = mean(total_weight_gain))
```


### Population numbers

#### Load the data <sup><font size="3em">[&#10059;]</font></sup>

```{r}
pop_data_file <- "https://raw.githubusercontent.com/MichielNoback/datasets/master/population/EDU_DEM_05022020113225751.csv"
local_file <- "population_data.csv"
if (! file.exists(local_file)) {
    download.file(pop_data_file, destfile = local_file)
}
```

```{r}
population <- read.table(local_file,
                         header = TRUE, 
                         sep=",")
```


#### Clean up <sup><font size="3em">[&#10059;&#10059;</font></sup>

**Which?**
`str(population)` tells me that `Unit.Code`, `Unit` and `PowerCode` are factors with one level. Using `table(population$PowerCode.Code, useNA = "always")` tells me that there are only zeros there. Same for `Reference.Period.Code` and `Reference.Period`. The variables `Flags` and `Flag.Codes` refer to the same, so one of them can be removed (I choose to remove `Flag.Codes`). The same counts for `SEX`/`Sex`, `AGE`/`Age` and `YEAR`/`Year`.

**Select**
This is my selection:

```{r results = 'hide'}
keep <- names(population)[c(1, 2, 4, 6, 8, 15, 16)]
keep
population <- as_tibble(population[, keep]) #tibble is nicer!
head(population)
```

#### Create a "wide" yearly report of totals <sup><font size="3em">[&#10059;&#10059;&#10059;</font></sup>


```{r wide-population-repost, eval = FALSE}
pop_totals <- dplyr::filter(population, Sex == "Total" & Age == "Total: All age groups")
##or, using base R
#population[population$Sex == "Total" & population$Age == "Total: All age groups", ]
pop_totals_wide <- pivot_wider(
    data = pop_totals[, c(1, 2, 5, 6)],
    names_from = Year,
    values_from = Value)
pop_totals_wide

```


#### Create a "wide" yearly report of population change <sup><font size="2em">[&#10059;&#10059;&#10059;&#10059;]</font></sup>

You could do this on the `pop_totals` dataset using the `lag()` function, after using `group_by()`, but since we have the wide format, you could also use a simple `for` loop, iterating the columns by index and subtracting the first from the second. I will demonstrate the only the first.

```{r pop-change-wide-report, eval = FALSE}
pop_totals %>% 
    group_by(Country) %>%
    mutate(Pop_change = as.integer(Value - lag(Value))) %>%
    ungroup() %>%
    select(-Value) %>%
    pivot_wider(names_from = Year,
        values_from = Pop_change) %>%
    select(-`2005`, -`2010`, -Flag.Codes) 
## backticks in above selection are required 
## because we are selecting names that are numbers!
```


#### Create a bar plot  <sup><font size="3em">[&#10059;&#10059;&#10059;</font></sup>

```{r pop-barplot, fig.asp=0.7, out.width='80%', fig.align='center', fig.cap="Barplot of population numbers in some European countries"}
sel <- population %>%
    filter(Sex == "Women" | Sex == "Men") %>%
    filter(COUNTRY %in% c("BEL", "CHE", "DNK", "FRA", "IRL", "DEU", "LUX", "NLD", "GBR")) %>%
    drop_na()

ggplot(sel, mapping = aes(Year)) +
    geom_bar(aes(weight = Value, fill = Sex))  + 
    facet_wrap(. ~ COUNTRY)

```


#### Highest growth rate <sup><font size="2em">[&#10059;&#10059;&#10059;&#10059;]</font></sup>

```{r pop-fastest-growth-rate, eval = FALSE}
population %>%
    filter(Sex == "Total" & Age == "Total: All age groups" & (Year == 2005 | Year == 2017)) %>%
    group_by(Country) %>%
    mutate(Change = Value - lag(Value), 
           Previous = lag(Value)) %>%
    mutate(GrowthRate = Change / Previous * 100) %>%
    ungroup() %>%
    select(COUNTRY, Country, GrowthRate ) %>%
    arrange(desc(GrowthRate)) %>%
    head(3)
```


### Rubella and Measles

No solutions (yet) for this exercise.




## The `lubridate` package


### Female births

#### Load the data <sup><font size="2em">[&#10059;&#10059;]</font></sup>

```{r results = 'hide'}
remote_file <- "https://raw.githubusercontent.com/MichielNoback/datasets/master/female_births/daily-total-female-births.csv"
local_file <- "../female_births.csv"

if (! file.exists(local_file)) {
    download.file(remote_file, destfile = local_file)
}
female_births <- read.table(local_file,
                          header = TRUE,
                          sep = ",",
                          stringsAsFactors = F)
female_births <- female_births %>% 
    mutate(Date = as_date(female_births$Date))

(female_births <- as_tibble(female_births))
```

Note that in this case `as_tibble(female_births)` with the original dataframe would have converted the Date column to date objects as well.

#### Check for missing rows <sup><font size="2em">[&#10059;&#10059;&#10059;]</font></sup>


```{r female-births-missing-data, eval = FALSE}
female_births %>%
    mutate(lagged = lag(Date),
           diff = as.integer(Date - lagged)) %>%
    filter(is.na(diff) | diff != 1)
```

#### Report birth numbers <sup><font size="2em">[&#10059;&#10059;&#10059;]</font></sup>

Report the number of births per month (Jan - Dec) and per weekday (Mon - Sun), as barplots.
Is there a month or day that seems to be anomalous, or do you see seasonal trends? Try a statistical test!

First create the required columns:

```{r female-births-numbers, eval = TRUE}
female_births <- female_births %>%
    mutate(Day = wday(Date, label = T),
           Month = month(Date, label = T))
```

Report monthly counts, as barplot.

```{r births-month-barplot, fig.asp=0.7, out.width='70%', fig.align='center', fig.cap="Monthly birth counts"}
#library(ggplot2)
female_births %>%
    ggplot(mapping = aes(x = Month, weight = Births)) +
    geom_bar(fill = "darkblue") +
    theme_bw()
```

The days

```{r births-month-barplot2, fig.asp=0.7, out.width='70%', fig.align='center', fig.cap="Daily birth counts"}
library(ggplot2)
female_births %>%
    ggplot(mapping = aes(x = Day, weight = Births)) +
    geom_bar(fill = "darkblue") +
    theme_bw()
```


### Diabetes

The `datasets` repo of this course contains a folder named `UCI_diabetes_dataset`. The complete dataset contains various types of information for 70 diabetes patients. Study the Readme file of this dataset before proceeding.

#### Create a codebook <sup><font size="2em">[&#10059;&#10059;]</font></sup>

Copy-and-paste the Code Fields into a file named `codebook.txt` and load it as a dataframe with name `codebook_diab`.

```{r load-codebook-diab}
codebook_diab <- read.table("data/diabetes_codebook.txt", sep = ",", header = TRUE)
codebook_diab <- as_tibble(codebook_diab)
head(codebook_diab)
```


#### Create a function to load patient data <sup><font size="2em">[&#10059;&#10059;&#10059;]</font></sup>

Create a function -`load_patient_data(<patient-number>)`- that can be used to load data for specific patients from file and return this as a dataframe. If you access the remote copies of the files you should cache the data locally so a second requests for the same data will return a local copy of the file. The function should give a friendly error message when there is no data for the specific patient (e.g. patient 99 is not present in the dataset). 

As extra challenge, you could read the documentation on the `tryCatch()` function and try to use this for your remote file access.


```{r function-to-load-patient-data}
load_patient_data <- function(patient_number) {
    base_url <-"https://raw.githubusercontent.com/MichielNoback/datasets/master/UCI_diabetes_dataset/data-"
    if (patient_number <= 9) patient_number <- paste0("0", patient_number)
    remote_file <- paste0(base_url, patient_number)
    local_file <- paste0("../diab-patient-", patient_number)
    column_names <- c("Date", "Time", "Code", "Value")
    #format:
    #04-21-1991	9:09	58	100
    result <- NULL
    if (! file.exists(local_file)) {
        result <- tryCatch( {#success
                message(paste0("downloading ", remote_file))
                download.file(remote_file, destfile = local_file)
                patient_data <- read.table(local_file, 
                                           sep = "\t", 
                                           stringsAsFactors = F)
                names(patient_data) <- column_names
                return(patient_data)
            },
            error = function(cond) {
                message(paste("Remote file does not seem to exist:", remote_file))
                message("Here's the original error message:")
                message(cond)
                return(NULL)
            },
            warning = function(cond) {
                message(paste("URL caused a warning:", remote_file))
                message("Here's the original warning message:")
                message(cond)
                return(NULL)
            })
    } else {
        message(paste0("reading local cache of", local_file))
        result <- read.table(local_file, 
                             sep = "\t",
                             stringsAsFactors = F)
        names(result) <- column_names
    }
    return(result)
}
patient8 <- load_patient_data(8)
head(patient8)
str(patient8)
```

B$






## The `stringr` package




